GPU Acceleration of the Material Point Method

### 2019-09-23 pres/pres.tex
redid some things

\section{Motivation on a GPU MPM Approach}

@@

\begin{frame}
PIC ideas:
\vfill
\begin{minipage}{0.45\paperwidth}
\begin{itemize}
    \item Combine Lagrangian particles \& Eulerian grid    \item Particles store all information
\end{itemize}
\vfill
Typical PIC/MPM roundtrip:
\vfill
\begin{enumerate}
  \item Particle-to-grid(P2G) transfer to an unmoving grid
    \item Solve discretized governing equations on grid
    \item Grid-to-particle(G2P) transfer back to particles \& move them
    \item[$\Rightarrow$] \textbf{meshfree, non-empirical}
\end{enumerate}
\end{minipage}
\begin{minipage}{0.35\paperwidth}
	\includesvg[width=0.35\paperwidth]{media/pic.svg}
	\tiny{Transfers: Interpolation functions are defined over grid nodes.}
\end{minipage}
\end{frame}
\subsection{GPGPU for performance enthusiasts}
\begin{frame}
  \frametitle{GPGPU for performance enthusiasts}
  Why would('nt) you?
  \vfill
  \begin{columns}[T]
    \begin{column}{0.45\textwidth}
      \textbf{Drawbacks:}
    \begin{itemize}
      \item Interactivity much easier on CPU, but slow PCI-Bus communication
      \item Code is mostly written against GPU architecture
      \item A lot of strain on the programmer
    \end{itemize}
    \end{column}
    \begin{column}{0.45\textwidth}
      \textbf{Benefits:}
    \begin{itemize}
      \item Data is already on the GPU for rendering
      \item \textbf{Higher parallelization acceleration}
    \end{itemize}
    \end{column}
  \end{columns}
\end{frame}
\section{A Gentle Introduction to the MPM}

@@

\subsection{The Pretty Strong but Mathematically Weak Formulation}
\begin{frame}
\frametitle{The Pretty Strong but Mathematically Weak Formulation}

\vfill
\begin{minipage}{0.49\textwidth}
  Heaviside step function has no weak derivate: \\\\
  $s= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
\end{minipage}
\begin{minipage}{0.4\textwidth}
\begin{tikzpicture}
  \small
  \begin{axis}[axis lines = middle, width=5.1cm,height=3.5cm,xtick={-1,0,1},ytick={0,1},ymin=-0.3,ymax=1.25,xmin=-1.8,xmax=1.8,ylabel=$s$]
                \addplot[blue,mark=*,mark options={fill=white},samples at={-2.1,0}] {0};
                \addplot[blue,mark=*,samples at={0,2.1}] {1.0};
            \end{axis}
\end{tikzpicture}
\end{minipage}
\vfill
Allows for point loads, material discontinuities and more. \cite{bathe2006finite}
\end{frame}
\subsection{Discretization of Space and Time}
\begin{frame}
\frametitle{Discretization of Space and Time}
\textbf{Time discretization} with implicit midpoint scheme:
  \begin{equation}
    \frac { y ^ { n + 1 } - y ^ { n } } { \Delta t } = f ^ { n + \frac{1}{2} } = f \left( t ^ { n } + \frac{\Delta t}{2}, \frac{1}{2} y ^ { n } + \frac{1}{2} y ^ { n + 1 } \right)
  \end{equation}
\begin{itemize}
  \item \textbf{implicit} requires linear system solve $\Rightarrow$ more stable, larger time steps
  \item \textbf{midpoint} as it conserves gov. equations
\end{itemize}

  \begin{equation}
    \Rightarrow \int _ { \Omega ^ { t } } _t\boldsymbol{q} \cdot {}_t\rho(_t\boldsymbol{v}^{n+1}-{}_t\boldsymbol{v}^{n}) d _t\boldsymbol{x}= \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}^{n+\frac{1}{2}}  d_t\boldsymbol{x}.
  \end{equation}
\end{frame}

\addplot [blue,thick] {func(x)};
\end{axis}
\end{tikzpicture}
\end{frame}
\begin{frame}
Thus set in:
$$
{}_t\boldsymbol{q}(\boldsymbol{x},t^n) = \sum_i \boldsymbol{e}_iw_i(\boldsymbol{x}),
{}_t\boldsymbol{v}^{n(+1)}(\boldsymbol{x}) = \sum_j \boldsymbol{v}^{n(+1)}_jw_j(\boldsymbol{x}).
$$
Combine it with numerical integration where the particles function as quadrature points \cite{Steffen}:

\begin{minipage}{0.6\textwidth}

$$
g _ { i } = \int _ { \Omega } g ( \boldsymbol{x} ) w _ { i } ( \boldsymbol{x} ) d \boldsymbol{x}
$$
\begin{equation}
  \approx \sum _ { p } g _ { p }  w_i(\boldsymbol{x}_p) V _ { p }.
\end{equation}
due to integration by midpoint rule:
\begin{equation}
\int _ { \Omega } f ( x ) d x \approx \sum _ { i = 1 } ^ { N } f \left( x _ { i } \right) h _ { i }.
\end{equation}

\end{minipage}
\begin{minipage}{0.3\textwidth}
  \begin{tikzpicture}[/pgf/declare function={f=-.1*(x)*(x-8);}]
\pgfplotsset{
    integral segments/.code={\pgfmathsetmacro\integralsegments{#1}},
    integral segments=3,
    integral/.style args={#1:#2}{
        ybar interval,
        domain=#1+((#2-#1)/\integralsegments)/2:#2+((#2-#1)/\integralsegments)/2,
        samples=\integralsegments+1,
        x filter/.code=\pgfmathparse{\pgfmathresult-((#2-#1)/\integralsegments)/2}
    }
}
\begin{axis}[
    ylabel=$f$,
    width =5cm,height=5cm,
    domain=0:6,
    samples=100,
    axis lines=middle
]
\addplot [thick] {f};
\addplot [
    red,
    integral=0:6
] {f};
\end{axis}
\end{tikzpicture}
\end{minipage}
\end{frame}
  \subsection{Velocity Fields: APIC-Transfers}
\begin{frame}
  \frametitle{Velocity Fields: APIC-Transfers}
  \begin{minipage}{0.64\textwidth}
  PIC-transfer
  \begin{enumerate}
    \item $(m\boldsymbol{v})_i^n = \sum_p w_{ip}^nm_p\boldsymbol{v}^n_p$
    \item $\boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}$
    \item $\boldsymbol{v}_{p,{PIC}}^{n+1} = \sum_i w^n_{ip}\boldsymbol{v}_i^{n+1}$
  \end{enumerate}
\vspace*{15}
APIC-transfers add a local velocity field $\boldsymbol{C}^n_p$ around $\boldsymbol{v}^n_p$:
\\\vspace*{-10}\\
$ ( m \boldsymbol{v} ) _ { i } ^ { n } = \sum _ { p } w _ { i p } ^ { n } m _ { p } \left( \boldsymbol{v} _ { p } ^ { n } + \boldsymbol{C} _ { p } ^ { n } ( \boldsymbol{x} _ { i } ^ { n } - \boldsymbol{x} _ { p } ^ { n }))$
\end{minipage}
\begin{minipage}{0.30\textwidth}
\begin{minipage}{1.0\textwidth}
  \small
  \includesvg[width=1.0\textwidth]{media/pic.svg}
  P2G-transfer $\sum_p$
\\
\end{minipage}
\begin{minipage}{1.0\textwidth}
  \small
  \includesvg[width=1.0\textwidth]{media/g2p.svg}
  G2P-transfer $\sum_i$
\end{minipage}
\end{minipage}
\end{frame}
\section{A MPM Guide on GPGPU Pitfalls and Optimizations}
\subsection{Layout of the data: SoA vs. AoS}
\begin{frame}[fragile]
\frametitle{Layout of the data: SoA vs. AoS}

@@

\begin{frame}
Sorting can dramatically increase workload performance of subsequent steps for neighboring queries:
\begin{enumerate}
  \item Deep sorted accesses are now \textbf{coalesced}. \label{it:sort_1}
  \item \textbf{Data reuse} due to L2-Cache and/or shared memory. \label{it:sort_2}
\end{enumerate}
\textbf{Double buffer particles} to use last sorted state as input for new sorting to profit from item 1 and 2!
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | r | r | r | r| r |r|}    \hline
    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
    Deep sorted           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
    \hline
 \end{tabular}
\caption{\footnotesize Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
\end{table}
\end{frame}
\subsection{The MPM Specific Transfers}
\begin{frame}
\frametitle{The MPM Specific Transfers}
All MPM operations belong to one of those parallelization schemes:
\begin{itemize}
 \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
 \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
\end{itemize}
\vfill
MPM-Transfers are executed \textbf{multiple times per physical frame} with varying numbers of variables and mathematical operations.\\\vspace{5}
$\Rightarrow$ Preprocessing steps only need to be done \textbf{once per physical frame}. Sorting already introduced as one of these.
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  G2P-Transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_inv_par}
  Pull-Operations read!
\end{minipage}
\end{frame}
\begin{frame}
Similar to filter or stencil operations on the GPU:
\begin{itemize}
  \item 1 node : 1 thread, split grid into blocks corr. to thread groups.
  \item Interpolation function however dependent on particle position.
  \item Needs to be rerun for every particle in the cell.
\end{itemize}
\vfill
Typical setup of transfers:
\begin{enumerate}
  \item Initialize shared memory (pull: with nodes from global memory).
  \item Perform transfers.
  \item Write back to global memory (push: global atomics since writes on halo).
\end{enumerate}
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_par}
  Pull-Operations read, Push-operations write!\\
  $\Rightarrow$ Race conditions.
\end{minipage}
\end{frame}
\begin{frame}
  Typically simulation domain(grid) much bigger than simulation model. Filtering inactive blocks as a preprocess improves performance.
\begin{itemize}
  \item Block is active if any cell counter is active.
  \item Cell counter is active if it has at least one particle.
\end{itemize}
\vfill
\begin{minipage}{0.9\textwidth}
  \footnotesize
  \includesvg[svgpath =../images/,width=\textwidth]{indirect_dispatch}
\end{minipage}
\end{frame}
\begin{frame}
\begin{itemize}
  \item Block and halo are always target of shared memory operations $\Rightarrow$ P2G-pull low occupancy.
  \item \textbf{Batching} multiple particles can increase performance due to hiding synchronization, unroll!
  \item Transfers respect shared memory bank conflicts fully.
  \item \textbf{Warp divergence} for varying cell counts.
\end{itemize}

@@

\selectlanguage{english}
\printbibliography
\end{document}

>>>>

\section{Motivation on a GPU MPM Approach}
\subsection{GPGPU for performance enthusiasts}
\begin{frame}
\frametitle{GPGPU for performance enthusiasts}
  Why would('nt) you?
  \vfill
  \begin{columns}[T]
    \begin{column}{0.45\textwidth}
      \textbf{Drawbacks:}
    \begin{itemize}
      \item Interactivity much easier on CPU, but slow PCI-Bus communication
      \item Code is mostly written against GPU architecture
      \item A lot of strain on the programmer
    \end{itemize}
    \end{column}
    \begin{column}{0.45\textwidth}
      \textbf{Benefits:}
    \begin{itemize}
      \item Data is already on the GPU for rendering
      \item \textbf{Higher parallelization acceleration}
    \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}
\begin{minipage}{0.45\paperwidth}
PIC ideas:
\begin{itemize}
    \item Particles store all information
    \item Eulerian grid as a non-deformable scratchpad
    \item[$\Rightarrow$] \textbf{meshfree}
\end{itemize}
\vfill
Typical PIC/MPM roundtrip:
\begin{enumerate}
  \item Transfer particle quantities to the grid (\textit{P2G})
  \item Solve discretized governing equations on grid
  \item Transfer back (\textit{G2P})
\end{enumerate}
\end{minipage}
\begin{minipage}{0.30\paperwidth}
\begin{minipage}{1.0\textwidth}
  \small
  \includesvg[width=1.0\textwidth]{media/pic.svg}
  P2G-transfer $\sum_p$
\\
\end{minipage}
\begin{minipage}{1.0\textwidth}
  \small
  \includesvg[width=1.0\textwidth]{media/g2p.svg}
  G2P-transfer $\sum_i$
\end{minipage}
\end{minipage}
\end{frame}
\section{A Gentle Introduction to the MPM}

@@

\subsection{Discretization of Space and Time}

\begin{frame}
\frametitle{Discretization of Space and Time}
\textbf{Space Discretization} is done in a Galerkin/FEM fashion with grid based interpolants $w_i$. Here dyadic products of cubic b-splines suffice. $w_i$ should satisfy at least \cite{gao2017adaptive}:
\begin{itemize}
  \item Partition of unity: $\sum_i w (\boldsymbol{x}-\boldsymbol{x}^n_i)=1$
  \item Identity relation: $\sum_i \boldsymbol{x}_i w (\boldsymbol{x}-\boldsymbol{x}^n_i)=\boldsymbol{x}$ $(\boldsymbol{x}=\boldsymbol{x}_p)$
  \item Non-negativity: $w \geq 0$.
  \item Limited support.
  \item $C^1$-continuity.
\end{itemize}
Shortening $w^n_{ip} = w(\boldsymbol{x}_p^n - \boldsymbol{x}_i^n).
\begin{tikzpicture}[remember picture,overlay,
  declare function={
    func(\x)= and(abs(\x) >= 0 , abs(\x)<1) * (pow(abs(\x),3)/2 - pow(abs(\x),2) +2/3)  +
	      and(abs(\x) >= 1 , abs(\x)<2) * (1/6* pow((2-abs(\x)),3))
   ;
  }
]
\begin{axis}[
  width=7.5cm,height=3.5cm,
  axis x line=middle, axis y line=middle,
  ymin=-0.1, ymax=1, ytick={0,1}, ylabel=$w$,
  xmin=-2.5, xmax=2.5, xtick={-2,...,2}, xlabel=$x$,
  domain=-2.5:2.5,samples=101, % added
]

\addplot [blue,thick] {func(x)};
\end{axis}
\end{tikzpicture}
\end{frame}

\begin{frame}
Transfer quantities from particles to grid. Numerical integration where the particles function as quadrature points \cite{Steffen}:

$$m_i = \int_\Omega \rho(\boldsymbol{x})w_i(\boldsymbol{x}) d\Omega \approx \sum_p \rho_p w_{ip} V_p \approx \sum_p m_p w_{ip}$$
$$A _ { i } = \int _ { \Omega }  A ( \boldsymbol{x} ) w _ { i } ( \boldsymbol{x} ) d \Omega \approx \sum _ { p } A _ { p }  w_{ip} V _ { p }.$$ \\
\begin{minipage}{0.62\textwidth}
APIC-transfers add a local velocity field $\boldsymbol{C}_p$ around $\boldsymbol{v}_p$:
\\\vspace*{-10}\\
$ ( m \boldsymbol{v} ) _ { i } = \sum _ { p } w _ { i p } m _ { p } \left( \boldsymbol{v} _ { p } + \boldsymbol{C} _ { p } ( \boldsymbol{x} _ { i } - \boldsymbol{x} _ { p } ))$
\end{minipage}
\begin{minipage}{0.33\textwidth}
\begin{minipage}{1.0\textwidth}
  \small
  \includesvg[width=1.0\textwidth]{media/pic.svg}
  P2G-transfer $\sum_p$
\\
\end{minipage}

\end{minipage}
\end{frame}

\begin{frame}
\begin{minipage}{0.64\textwidth}
The MPM can be described in the weak formulation (FEM: virtual quantities). This allows us to 'shift' the derivative:\begin{itemize}
\item $\boldsymbol{A}_p = \sum_i \boldsymbol{A}_i w_{ip} $
\item $\nabla \boldsymbol{A}_p = \sum_i \boldsymbol{A}_i \nabla w_{ip}^T $
\item$\nabla \cdot \boldsymbol{A}_p = \sum_i \boldsymbol{A}^T_i \nabla w_{ip} $
\end{itemize}
\end{minipage}
\begin{minipage}{0.30\textwidth}
\begin{minipage}{1.0\textwidth}
  \small
  \includesvg[width=1.0\textwidth]{media/g2p.svg}
  G2P-transfer $\sum_i$
\end{minipage}
\end{minipage}
\vfill
\textbf{Time discretization} with implicit midpoint scheme:\\
$v_i ^ { n + 1 } - v_i ^ { n } = \frac{\Delta t}{m_i^n}f_i ^ { n + \frac{1}{2} } =\frac{\Delta t}{m_i^n} f \left( t ^ { n } + \frac{\Delta t}{2}, \frac{1}{2} y ^ { n } + \frac{1}{2} y ^ { n + 1 } \right)$
\begin{itemize}
  \item \textbf{implicit} requires linear system solve $\Rightarrow$ more stable, larger time steps
  \item \textbf{midpoint} as it conserves angular momentum
\end{itemize}
\end{frame}
\section{A MPM Guide on GPGPU Pitfalls and Optimizations}
\subsection{Layout of the data: SoA vs. AoS}
\begin{frame}[fragile]
\frametitle{Layout of the data: SoA vs. AoS}

@@

\begin{frame}
Sorting can dramatically increase workload performance of subsequent steps for neighboring queries:
\begin{enumerate}
  \item Deep sorted accesses are now \textbf{coalesced}. \label{it:sort_1}
  \item \textbf{Data reuse} due to L2-Cache and/or shared memory. \label{it:sort_2}
\end{enumerate}
\textbf{Double buffer particles} to use last sorted state as input for new sorting to profit from item 1 and 2!
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | r | r | r | r| r |r|}    \hline
    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
    Deep sorted           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
    \hline
 \end{tabular}
\caption{\footnotesize Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
\end{table}
\end{frame}
\subsection{The MPM Specific Transfers}
\begin{frame}
\frametitle{The MPM Specific Transfers}
All MPM operations belong to one of those parallelization schemes:
\begin{itemize}
 \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item 1 thread : 1 node: $\square_i = \square_i \circ \square_i \circ ... \circ \square_i$.
 \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
 \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
\end{itemize}
\vfill
MPM-Transfers are executed \textbf{multiple times per physical frame} with varying numbers of variables and mathematical operations.\\\vspace{5}
$\Rightarrow$ Preprocessing steps only need to be done \textbf{once per physical frame}. Sorting already introduced as one of these.
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  G2P-Transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_inv_par}
  Pull-Operations read!
\end{minipage}
\end{frame}
\begin{frame}
Similar to filter or stencil operations on the GPU:
\begin{itemize}
  \item 1 node : 1 thread, split grid into blocks corr. to thread groups.
  \item Interpolation function however dependent on particle position.
  \item Needs to be rerun for every particle in the cell.
\end{itemize}
\vfill
Typical setup of transfers:
\begin{enumerate}
  \item Initialize shared memory (pull: with nodes from global memory).
  \item Perform transfers.
  \item Write back to global memory (push: global atomics since writes on halo).
\end{enumerate}
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_par}
  Pull-Operations read, Push-operations write!\\
  $\Rightarrow$ Race conditions.
\end{minipage}
\end{frame}
\begin{frame}
  Typically simulation domain(grid) much bigger than simulation model. Filtering inactive blocks as a preprocess improves performance.
\begin{itemize}
  \item Block is active if any cell counter is active.
  \item Cell counter is active if it has at least one particle.
\end{itemize}
\vfill
\begin{minipage}{0.9\textwidth}
  \footnotesize
  \includesvg[svgpath =../images/,width=\textwidth]{indirect_dispatch}
\end{minipage}
\end{frame}
\begin{frame}
\begin{itemize}
  \item Block and halo are always target of shared memory operations $\Rightarrow$ P2G-pull low occupancy.
  \item \textbf{Batching} multiple particles can increase performance due to hiding synchronization, unroll!
  \item Transfers respect shared memory bank conflicts fully.
  \item \textbf{Warp divergence} for varying cell counts.
\end{itemize}

@@

\selectlanguage{english}
\printbibliography

\begin{frame}
\frametitle{The Pretty Strong but Mathematically Weak Formulation}

\vfill
\begin{minipage}{0.49\textwidth}
  Heaviside step function has no weak derivate: \\\\
  $s= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
\end{minipage}
\begin{minipage}{0.4\textwidth}
\begin{tikzpicture}
  \small
  \begin{axis}[axis lines = middle, width=5.1cm,height=3.5cm,xtick={-1,0,1},ytick={0,1},ymin=-0.3,ymax=1.25,xmin=-1.8,xmax=1.8,ylabel=$s$]
                \addplot[blue,mark=*,mark options={fill=white},samples at={-2.1,0}] {0};
                \addplot[blue,mark=*,samples at={0,2.1}] {1.0};
            \end{axis}
\end{tikzpicture}
\end{minipage}
\vfill
Allows for point loads, material discontinuities and more. \cite{bathe2006finite}
\end{frame}

\addplot [blue,thick] {func(x)};
\end{axis}
\end{tikzpicture}
\end{frame}
\section*{Rest}
\end{document}
